### 2024-11-23 Article/Background_sections/Limit.tex
final

The following lemma is a classical result, its proof is in the appendix.

@@

@@

The connection between constructions as products and equalizers and limits is made clear by the following theorem. The idea behind the proof is the fact that, given a diagram $D : \cat I \rightarrow \cat C$, if each subset of objects $X = \{D(i) \mid i \in \Ob(\cat I)\} \subseteq \Ob(\cat C)$ has a product $(\prod_{i \in I} D(i), (\pi_i)_{i \in \Ob( \cat I)})$ and each pair of arrows $f, g \in \cat C (D(i), D(j))$ has an equalizer, then one can construct the cone taking the equalizer of the arrows that has as domain the product of the objects of the diagram, and as codomain the product of the codomains of the arrows of the diagram. This construction has the universal property because equalizers and products do. A detailed proof is in the appendix.

@@

>>>>

The following lemma is a classical result.

@@

\begin{proof}
    For the ``only if'' part, suppose the left square to be a pullback. To verify the outer rectangle is a pullback, consider the following situation:
    \[
        \begin{tikzcd}[row sep = 20, column sep = 20]
            P \arrow[bend left=30]{rrrd}{p_1} \arrow[bend right=30, swap]{ddr}{p_2} & & \\
            & & & Z \arrow[d, "h"] \\
            & A \arrow[r, "a"swap] & B \arrow[r, "b"swap] & C
        \end{tikzcd}
    \]
    But the right square is a pullback implies that there exists a unique $u$ such that
    \[
        \begin{tikzcd}[row sep = 20, column sep = 20]
            P \arrow[bend left=30]{rrrd}{p_1} \arrow[bend right=30, swap]{ddr}{p_2} \arrow[bend left=15, dashed]{drr}{u}& & \\
            & &  Y \arrow[r, "g"] \arrow[d, "k"]& Z \arrow[d, "h"] \\
            & A \arrow[r, "a"swap] & B \arrow[r, "b"swap] & C
        \end{tikzcd}
    \]
    And, since the left square is a pullback, there exists a unique $v$ such that
    \[
        \begin{tikzcd}[row sep = 20, column sep = 20]
            P \arrow[bend left=30]{rrrd}{p_1} \arrow[bend right=30, swap]{ddr}{p_2} \arrow[bend left=15]{drr}{u} \arrow[dr, "v", dashed]& & \\
            & X \arrow[r, "f"] \arrow[d, "t"swap] &  Y \arrow[r, "g"] \arrow[d, "k"]& Z \arrow[d, "h"] \\
            & A \arrow[r, "a"swap] & B \arrow[r, "b"swap] & C
        \end{tikzcd}
    \]
    Hence, the whole rectangle is a pullback.

    For the ``if'' part, consider the following situation.
    \[
        \begin{tikzcd}[row sep = 20, column sep = 20]
            P \arrow[bend left=30]{rrd}{p_1} \arrow[bend right=30, swap]{ddr}{p_2} & \\
            & & Y \arrow[d, "k"] \\
            & B \arrow[r, "b"swap] & C
        \end{tikzcd}
    \]
    We have now to show that the unique arrow $v: P \to X$ (of the outer rectangle) is such that $f \circ v = p_1$, but this follows from the fact that the right square is a pullback, having the following situation.
    \[
        \begin{tikzcd}[row sep = 20, column sep = 20]
            P \arrow[bend left=30]{rrrd}{g \circ p_1} \arrow[bend right=30, swap]{ddr}{p_2} \arrow[bend left=15, dashed]{drr}{u}& & \\
            & &  Y \arrow[r, "g"] \arrow[d, "k"]& Z \arrow[d, "h"] \\
            & A \arrow[r, "a"swap] & B \arrow[r, "b"swap] & C
        \end{tikzcd}
    \]
\end{proof}

@@

The connection between constructions as products and equalizers and limits is made clear by the following theorem. The idea behind the proof is the fact that, given a diagram $D : \cat I \rightarrow \cat C$, if each subset of objects $X = \{D(i) \mid i \in \Ob(\cat I)\} \subseteq \Ob(\cat C)$ has a product $(\prod_{i \in I} D(i), (\pi_i)_{i \in \Ob( \cat I)})$ and each pair of arrows $f, g \in \cat C (D(i), D(j))$ has an equalizer, then one can construct the cone taking the equalizer of the arrows that has as domain the product of the objects of the diagram, and as codomain the product of the codomains of the arrows of the diagram. This construction has the universal property because equalizers and products do.

@@

\begin{proof}
    Let $D: \cat{I \rightarrow C}$ a diagram, with $\cat I$ finite.

    The \emph{if} statement follows from definitions of products and equalizers (\Cref{def:pullback_pushout}, \Cref{def:equalizer_coequalizer})

    To satisfy the \emph{only if} statement, we want an object $L$ together with morphisms $p_i : L \rightarrow D(j)$ such that:
    \begin{enumerate}
        \item\label{item:cone} $\{p_i: L \rightarrow D(i)\}$ is a cone -- i.e., for each morphism of $\cat I$ $\alpha : i \rightarrow j$, $D(\alpha) \circ p_i = p_j$; and
        \item\label{item:univ_prop} for each $E$ and $q_i : E \rightarrow D(j)$ in $\cat C$, with $D(\alpha) \circ q_i = q_j$ for each $\alpha : i \rightarrow j$ of $\cat I$, there exists a unique $f: E \rightarrow L$ such that $q_i = p_i \circ f$ for each $i \in \Ob(\cat I)$.
    \end{enumerate}

    Consider the two products (which exist by hypothesis) $\prod_{j \in \Ob(\cat I)} D(j)$, the product of the objects of the diagram, and $\prod_{\alpha \in \Hom(\cat I)}D(cod \ \alpha)$, the product of the codomains of the morphisms in $D$, where $\pi_x$ is the $x$-th projection of the product.
    Let now:
    \[
        \gamma, \varepsilon : \prod_{j \in \Ob(\cat I)}D(j) \ \longrightarrow \prod_{\alpha \in \Hom(\cat I)} D(cod \ \alpha)
    \]
    be defined by $\gamma_\alpha = \pi_{D(cod\ \alpha)}$ (the projection on the codomain of $\alpha$) and $\varepsilon_\alpha = D(\alpha) \circ \pi_{D(dom \ \alpha)}$.
    Let now $e: L \rightarrow \prod_{j \in \Ob(\cat I)}D(j)$ the equalizer of $\gamma$ and $\varepsilon$ (which exists by hypothesis), and, for each $j \in \Ob(\cat I)$, $p_j: L \rightarrow D(j)$, defined by $p_j = \pi_{D(j)} \circ e$.

    What we want now is to show that $(L, (p_i))_{i \in \cat I}$ is the limit of $D$, namely, to prove that the conditions given at the beginning are valid.

    For condition~\ref{item:cone}, we have to show that, for each $\alpha : i \rightarrow j$ of $\cat I$, we have $D(\alpha) \circ p_i = p_j$:
    \begin{align*}
        D(\alpha) \circ p_i 
            &= D(\alpha) \circ \pi_{D(i)} \circ e   && \textit{Definition of $p_j$} \\
            &= \varepsilon_{\alpha} \circ e         && \textit{Definition of $\varepsilon$}\\
            &= \gamma_\alpha \circ e                && \textit{$e$ is an equalizer of $\pi, \varepsilon$}\\
            &= \pi_{D(j)} \circ e                   && \textit{Definition of $\pi$} \\
            &= p_j                                  && \textit{Definition of $p_j$} \\
    \end{align*}

    For condition~\ref{item:univ_prop}, suppose that $(E, (q_i)_{i \in \Ob(\cat I)})$ has the properties stated. By definition of product, there exists a (unique) arrow $q: E \rightarrow \prod_{j \in \Ob(\cat I)}D(j)$. For each arrow $\alpha: i \rightarrow j$, we have:
    \begin{align*}
            \gamma_\alpha \circ q 
                &= \pi_{D(j)} \circ q           && \textit{Definition of $\pi$} \\
                &= q_j                          && \textit{Definition of $q_j$} \\
                &= D(\alpha) \circ q_i          && \textit{Assumption on $q_j$} \\
                &= D(\alpha) \circ \pi_{D(j)} \circ q   &&\textit{Definition of $q_i$} \\
                &= \varepsilon_{\alpha} \circ q && \textit{Definition of $\varepsilon$} \\
    \end{align*}
    Since $e$ equalizes $\pi$ and $\varepsilon$, there exists a unique $f: E \rightarrow L$ in $\cat C$ such that $q = e \circ f$. Then, for each $j \in \Ob(\cat I)$, we have $\pi_{D(j)} \circ q = \pi_{D(j)} \circ e \circ f$, hence, $q_i = p_i \circ f$.
\end{proof}

### 2024-11-23 Article/introduction.tex
final

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}

\iffalse

@@

%The development of visual formalisms such as equality graphs (e-graphs) has to be brought back to the 1980s, in the context of automatic theorem provers (ATPs).
%^E-Graphs are an extension of prior structures, in order to represent in a more compact way equivalence classes of expressions, maintaining closure-under-congruence invariant.
%Graph manipulation turns out to be useful in practical fields such as systems optimisation or components upgrades.
%The development of formal frameworks is therefore necessary, especially for what concerns parallelism and concurrency, as the research line in which this work wants to be integrated.

The development of the visual formalism of \emph{equality graphs} (e-graphs) can be traced back to the 1980s, originating in the context of automatic theorem provers (ATPs)\todo{mettere referenza}. E-graphs provide a compact and efficient way to represent equivalence classes of expressions while maintaining the crucial property of closure under congruence. This ensures that equivalences are preserved across all contexts in which an expression might appear, making e-graphs a robust tools for reasoning about equivalence of algebraic expressions.
%\todo{Non sicuro dell'ultima riga}.

Dependency and concurrency harden the process of program optimisation, because application order of rewrites can significantly improve the quality of the overall program.
Such problem is known as \textit{phase-ordering problem}, and a recent approach, known as equality saturation, was developed in \cite{Tate_2011}.
This technique provides an alternative to heuristic methods, consisting of keeping track of a set of candidate equivalent optimized programs, 
computed by repeatedly inferring equivalences between program fragments.
E-Graphs provide a compact representation of this set, making operation feasible in concrete applications.
%\todo{Non ho capito questo paragrafo}

The utility of e-graphs thus extends beyond theoretical applications, finding practical relevance in fields such as system optimisation and component upgrading.
%\todo{Qua servirebbero degli esempi di uso}
%
However, the more a formalism is used, the greater is the need for formal frameworks to guide its development and application. In particular, addressing issues of parallelism and concurrency is critical for 
a mathematically 
sound understanding of these structures.

Among the rewriting paradigms for graph-like structures, one of the leading proposal is the Double Pushout (DPO) approach \cite{Ehrig_Golas,ehrig2006fundamentals}, since it provides, paired with the framework of $\mathcal{M}$-adhesive  categories \cite{lack2005adhesive,azzi2019essence,behr2022fundamentals}, a robust mathematical foundation for the definition and the application of rules. 
%
Integrating the theory of e-graphs manipulation into the broader subject of DPO rewriting would put on firmer ground the status of e-graphs. Moreover, it would allow to apply to this new context the well known results about concurrency and parallelism already known provided by the abstract theory of DPO rewriting systems \cite{baldan1999concurrent,ehrig1999handbook}.

Such an integration is the main aim of the thesis.
 On the technical level, to reach our aim we need to prove some adhesivity property of e-graphs. 
 In order to do so, some consideration on graphs with equivalences are in order. 
 A graph with equivalence is nothing but a graph endowed with an equivalence relation among its nodes. 
 The main result on this kind of graphs is a weak adhesivity property that holds for regular monomorphisms. 
% Such result is, then, inherited by e-graphs \todo{Da qualche parte mettere riferimento a \cite{egg}}, which are graphs with equivalence which satisfies additional requirements that can be intuitively summarized as the property of edges to have equivalent sources whenever targets are equivalent.
%
%%On the technical level, to reach our objective we need to prove that the category of e-graphs exhibits a specific adhesivity property. To this end, it is necessary to first examine \emph{graphs with equivalence}, as they form the foundational structure underlying e-graphs.
%
%A graph with equivalence is simply a graph where an equivalence relation is defined on its nodes. The primary result concerning graphs with equivalences is an adhesivity property with respect to regular monomorphisms. 
%
This adhesivity result, in turn, extends naturally, via the results of \cite{castelnovo2022newcriterionmathcalmmathcalnadhesivity}, to e-graphs \todo{Add reference to \cite{egg} da qualche parte}, which in fact can be interpreted as a specialised class of graphs with equivalences that satisfy additional structural constraints. Specifically, e-graphs enforce a condition that edges must respect equivalence relations: if two edges have equivalent target nodes, then their source nodes must also be equivalent.

>>>>

\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
\markboth{INTRODUCTION}{}
\iffalse

@@

%The development of visual formalisms such as equality graphs (e-graphs) has to be brought back to the 1980s, in the context of automatic theorem provers (ATPs).
%^E-Graphs are an extension of prior structures, in order to represent in a more compact way equivalence classes of expressions, maintaining closure-under-congruence invariant.
%Graph manipulation turns out to be useful in practical fields such as systems optimization or components upgrades.
%The development of formal frameworks is therefore necessary, especially for what concerns parallelism and concurrency, as the research line in which this work wants to be integrated.

The development of the visual formalism of \emph{equality graphs} (e-graphs) can be traced back to the 1980s, originating in the context of automatic theorem provers (ATPs) \cite{Nelson1980}. 
E-graphs provide a compact and efficient way to represent equivalence classes of expressions while maintaining the crucial property of closure under congruence. 
This ensures that equivalences are preserved across all contexts in which an expression might appear, making e-graphs a robust tools for reasoning about equivalence of algebraic expressions.
%\todo{Non sicuro dell'ultima riga}.

Dependency and concurrency harden the process of program optimization, because application order of rewrites can significantly improve the quality of the overall program.
Such problem is known as \textit{phase-ordering problem}, and a recent approach, known as equality saturation, was developed in \cite{Tate_2011}.
This technique provides an alternative to heuristic methods, consisting of keeping track of a set of candidate equivalent optimized programs, 
computed by repeatedly inferring equivalences between program fragments.
E-Graphs provide a compact representation of this set, making operation feasible in concrete applications.
%\todo{Non ho capito questo paragrafo}

The utility of e-graphs thus extends beyond theoretical applications, finding practical relevance in fields such as system optimization and component upgrading. Example of uses can be found in \cite{Cheli2021}.
%\todo{Qua servirebbero degli esempi di uso}
%
However, the more a formalism is used, the greater is the need for formal frameworks to guide its development and application. In particular, addressing issues of parallelism and concurrency is critical for 
a mathematically 
sound understanding of these structures.

%Among the rewriting paradigms for graph-like structures, one of the leading proposal is the Double Pushout (DPO) approach \cite{Ehrig_Golas,ehrig2006fundamentals}, since it provides, paired with the framework of $\mathcal{M}$-adhesive  categories \cite{lack2005adhesive,azzi2019essence,behr2022fundamentals}, a robust mathematical foundation for the definition and the application of rules. 
Among the rewriting paradigms for graph-like structures, one of the leading proposal is the Double Pushout (DPO) approach \cite{Ehrig_Golas,ehrig2006fundamentals}, since it provides, paired with the framework of $\mathcal{M}$-adhesive  categories \cite{lack2005adhesive,azzi2019essence,behr2022fundamentals}, a robust mathematical foundation for the definition and the application of rules,
that us the identification of a subgraph to be removed and the insertion of a new subgraph.
%
Integrating the theory of e-graphs manipulation into the broader subject of DPO rewriting would put on firmer ground the status of e-graphs. Moreover, it would allow to apply to this new context the well known results about concurrency and parallelism already known provided by the abstract theory of DPO rewriting systems \cite{baldan1999concurrent,ehrig1999handbook}.

Such an integration is the main aim of the thesis.
 On the technical level, to reach our aim we need to prove some adhesivity property of e-graphs. 
 In order to do so, some consideration on graphs with equivalences are in order. 
 A graph with equivalence is nothing but a graph endowed with an equivalence relation among its nodes. 
 The main result on this kind of graphs is a weak adhesivity property that holds for regular monomorphisms. 
% Such result is, then, inherited by e-graphs \todo{Da qualche parte mettere riferimento a \cite{egg}}, which are graphs with equivalence which satisfies additional requirements that can be intuitively summarized as the property of edges to have equivalent sources whenever targets are equivalent.
%
%%On the technical level, to reach our objective we need to prove that the category of e-graphs exhibits a specific adhesivity property. To this end, it is necessary to first examine \emph{graphs with equivalence}, as they form the foundational structure underlying e-graphs.
%
%A graph with equivalence is simply a graph where an equivalence relation is defined on its nodes. The primary result concerning graphs with equivalences is an adhesivity property with respect to regular monomorphisms. 
%
 This adhesivity result, in turn, extends naturally, via the results of \cite{castelnovo2022newcriterionmathcalmmathcalnadhesivity}, to e-graphs, which in fact can be interpreted as a specialised class of graphs with equivalences that satisfy additional structural constraints. 
 An operational presentation of such structures can be found in \cite{egg}. 
 Specifically, e-graphs enforce a condition that edges must respect equivalence relations: if two edges have equivalent target nodes, then their source nodes must also be equivalent.
